---
title: "Data overview"
author: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  data_collection: "UUID"
  output_collection: "UUID"
  .arv_save:
    value:
      collection: "UUID"
      filename: "{basename(html_filename)}"
---

```{r libs, message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer )
library(gridExtra)
library(kableExtra)
library(DT)

```

```{r random seed}
set.seed(42)

addline_format <- function(x,...){
    gsub('\\s','\n',x)
}

```

Load ExpressionSet object.
```{r load eset, message=FALSE, warning=FALSE, results='asis'}

wd=getwd()
eset <- aws.s3::s3readRDS("tpm_eset.Rds", bucket = params$output_collection)

# removing the samples that do not have immune phenotype, that are part of the full dataset. 

cat("Data without immune phenotype is excluded (here as FALSE):")
kable( table(is.na(eset$CD8IMMPH)))

eset <- eset[, !is.na(eset$CD8IMMPH)]

  ### Prepping ematrix and annotation for
  TPM = Biobase::exprs(eset)
  pseudo=0.01
  ematrix=data.matrix(log2(TPM+pseudo))

  Design = Biobase::pData(eset) #sampleAnnot_complete.dt 
  Annot = Biobase::fData(eset)
  rownames(ematrix)=Annot$symbol; rm(TPM)
  
  # cleaning information 
  # UROTHELIAL BLADDER CANCER	 == gRED
  # BLADDER = pRED 
  
  # NSCLC = gRED
  # LUNG = pRED

  Design$INDICAT [ grep("BLADDER", Design$INDICAT) ] <- "BLADDER"

  # Removing gRED studies
  Design = Design[ ! Design$STUDYID %in% c("GO29293","GO28915","GO29294"), ]   # removed a total of 1395 samples

  
cat("Studies from gRED were removed: GO29293, GO28915,GO29294, to only contain new samples in the description area")
cat("Total samples included")
cat( nrow(Design))
  
# Including identifiers for full atezo annotation 
Design <- Design %>%
  rownames_to_column("A_SampleID")
  
# full information added here as extra_samples_annot, using rtbeda
full_sample_anno <- aws.s3::s3read_using(
  read_tsv,
  show_col_types = FALSE,
  object = "full_samples_anno.tsv",
  bucket = params$data_collection
)

extra_samples_anno <- aws.s3::s3read_using(
  read_tsv,
  object = "extra_samples_anno.tsv",
  bucket = params$data_collection
)

extra_samples_anno <- extra_samples_anno %>% filter( A_SampleID %in% Design$A_SampleID)

#### Replace indication with with curated annotations
ah_anno = read.table("../data/Indications_final.csv", sep=',',header=TRUE)
tmpid=paste(sapply(full_sample_anno$USUBJID,function(x) strsplit(x,"-")[[1]][1]),sapply(full_sample_anno$USUBJID,function(x) strsplit(x,"-")[[1]][3]), sapply(full_sample_anno$SampleID,function(x) strsplit(x,"_")[[1]][2]),sep="_")

tmpid2=paste(sapply(full_sample_anno$USUBJID,function(x) strsplit(x,"-")[[1]][1]),sapply(full_sample_anno$USUBJID,function(x) strsplit(x,"-")[[1]][3]),sep="_")

full_sample_anno_ah=full_sample_anno
full_sample_anno_ah$HA_Indication_Full=ah_anno[match(tmpid2,ah_anno$Study_Patient),"TumorDesc_Primary.Tumor.Diagnosis.Full.Text"]
full_sample_anno_ah$HA_Indication=ah_anno[match(tmpid2,ah_anno$Study_Patient),"TumorDesc_Primary.Tumor.Diagnosis"]
full_sample_anno_ah=full_sample_anno_ah[which(full_sample_anno_ah$A_SampleID%in%Design$A_SampleID),]

#indcheck=table(paste(full_sample_anno_ah$INDICAT, full_sample_anno_ah$HA_Indication, sep="_"))
#barplot(sort(indcheck,decreasing=TRUE)[1:40], las=2, cex.names=0.5)

```

# General information 

```{r general stats, message=FALSE, warning=FALSE, results='asis'}

cat("AGE, median -->")
summary(extra_samples_anno$AGE)["Median"]

cat("SEX -->")
kable(table(extra_samples_anno$SEX))


cat("VISIT_NAME -->")
kable(table(extra_samples_anno$VISIT_NAME))

cat("APCPI -->")
kable(table(extra_samples_anno$APCPI))

svg(paste(wd, "barplot_CPI_HA.svg", sep="/"), width = 3, height = 5)
cpi=data.frame(CPI=names(table(extra_samples_anno$APCPI)),Nr=c(table(extra_samples_anno$APCPI)))
ggplot(cpi, aes(x = "", y = Nr, fill=CPI)) +
  geom_col()  
dev.off()

Design_full <- left_join(Design, 
                    extra_samples_anno, 
                    by="A_SampleID")

Design_full <- left_join(Design_full, 
                    full_sample_anno_ah[,c("A_SampleID","HA_Indication")], 
                    by="A_SampleID")

Design <- left_join(Design, 
                    full_sample_anno_ah[,c("A_SampleID","HA_Indication")], 
                    by="A_SampleID")


### Take over the indications from curated table 
Design_full$INDICAT=Design_full$HA_Indication

Design$INDICAT=Design$HA_Indication
write.table(Design,"../data/Design_indications_final.tsv", sep="\t",row.names=TRUE, col.names=TRUE, quote=FALSE)


cat("Indicat -->")
kable(table(Design_full$INDICAT))

kable(Design_full %>% group_by( INDICAT, APCPI) %>% tally())

kable(Design_full %>% group_by( HA_Indication, APCPI) %>% tally())

#cat("LPCPID")
#kable(table(extra_samples_anno$LPCPID))

cat("LAPCPIN -->")
kable(table(extra_samples_anno$LAPCPIN ))

cat("DSTAGE (only indications with >5 entries per stage are displayed) -->")
kable(Design_full %>% group_by( INDICAT, DSTAGE) %>% tally() %>% filter(n>5))

#cat("Types of Lung cancer -->")
#kable(table(Design %>% filter(INDICAT == "LUNG") %>% pull (HIST)))

```
# Overall indication distribution 

* Indications with less than 5 samples not shown. Total samples `r nrow(Design)`

```{r pies 1, fig.height=6, fig.width=8, fig.align="center", message=FALSE, warning=FALSE}

tally_i = Design %>% group_by( INDICAT) %>% tally()
pie_data_descend <- tally_i %>% filter (n > 5) %>% drop_na(INDICAT)
#pie_data_descend <- pie_data_descend %>% arrange(-n)
pie_data_descend$INDICAT <- ordered(pie_data_descend$INDICAT)
pie_data_descend$INDICAT  <- ordered(pie_data_descend$INDICAT, levels = pie_data_descend %>% arrange(n) %>%pull(INDICAT))


col_indicat = colorRampPalette(brewer.pal(8,"Dark2"))(length(unique(pie_data_descend$INDICAT)))
names(col_indicat) = unique(pie_data_descend$INDICAT)     

p1= ggplot(pie_data_descend, aes(x= INDICAT, y=n)) + 
    geom_bar(stat="identity") + 
    ggtitle("n samples per cancer type") +
    coord_flip() +
    scale_fill_manual(values=col_indicat )+
    ylab("Number of samples") + xlab("Cancer type") + 
    theme_bw()
   
p1 

wd=getwd()
svg(paste(wd, "barplot_indication_HA.svg", sep="/"), width = 8, height = 5)
p1
dev.off()


tally_i = Design %>% group_by( INDICAT, CD8IMMPH) %>% tally()
tally_i_Indlevel = Design %>% group_by( INDICAT) %>% tally()
tally_i_Indlevel <- tally_i_Indlevel %>% filter (n > 5) %>% drop_na(INDICAT)

## sort this part!
tally_i_Indlevel2 <- tally_i_Indlevel %>% filter (n <= 5) %>% drop_na(INDICAT)

#pie_data_descend <- tally_i %>% filter (n > 4) %>% drop_na(INDICAT)
pie_data_descend <- tally_i %>% filter (INDICAT %in% tally_i_Indlevel$INDICAT) 

pheno_col <- c(
  INFLAMED = "#d95f02",
  EXCLUDED = "#1b9e77",
  DESERT = "#7570b3"
)

label_x = tally_i %>% group_by(INDICAT) %>% add_count(INDICAT, wt = n) %>% dplyr::select(INDICAT, nn) %>% distinct()
label_x = label_x[ match( unique(pie_data_descend$INDICAT), label_x$INDICAT), ]
label_x <- na.omit(label_x)
label_x <- label_x %>% arrange(nn)

pie_data_descend$INDICAT <- ordered(pie_data_descend$INDICAT)
pie_data_descend$INDICAT  <- ordered(pie_data_descend$INDICAT, levels = label_x$INDICAT)

label_x = paste(label_x$INDICAT, " (", label_x$nn, ")", sep="")

p1= ggplot(pie_data_descend, aes(x=INDICAT, y=n, fill=CD8IMMPH)) + 
    geom_bar(stat="identity") + 
    ggtitle("n samples per cancer type") +
    coord_flip() +
    scale_fill_manual(values=pheno_col) +
    ylab("Number of samples") + xlab("Cancer Type")+
    scale_x_discrete(labels=addline_format(label_x) ) + 
    theme_bw()
   

p1

svg(paste(wd, "barplot_indication_CD8IMMPH_HA.svg", sep="/"), width = 8, height = 6)
p1
dev.off()
 

```

# Indication by location 

Sorted by total freq: liver, lung, lymph node. Bars alphabetically ordered.

```{r pies 2, fig.height=5, fig.width=20, fig.align="center"}

# minimal sample size
n_min = 5

Design_i <- dplyr::filter(Design, grepl("ung",BMFLOC, ignore.case = TRUE))

tally_i = Design_i %>% group_by( INDICAT) %>% tally()
tally_i$perc <- round((tally_i$n*100)/sum(tally_i$n), digits = 1)
pie_data_descend <- tally_i %>% filter (n > n_min) %>% drop_na(INDICAT)

label_x = paste(pie_data_descend$INDICAT, " (", pie_data_descend$perc, "%)", sep="")

p1 = ggplot(pie_data_descend, aes(x= INDICAT, y=n, fill=INDICAT)) + 
    geom_bar(stat="identity") + 
    ggtitle(paste("Lung (n=", sum(pie_data_descend$n), ")", sep="" )) +
    coord_flip() +
    scale_fill_manual(values=col_indicat[pie_data_descend$INDICAT], drop=FALSE )+
    ylab("Number of samples") + xlab("Indication") + labs(caption = paste("", sep="")) +
    theme(legend.position="none") +
    scale_x_discrete(labels=addline_format(label_x) )+ 
    theme_bw()


####

Design_i <- dplyr::filter(Design, grepl("iver",BMFLOC, ignore.case = TRUE))

tally_i = Design_i %>% group_by( INDICAT) %>% tally()
tally_i$perc <- round((tally_i$n*100)/sum(tally_i$n), digits = 1)
pie_data_descend <- tally_i %>% filter (n > n_min) %>% drop_na(INDICAT)
label_x = paste(pie_data_descend$INDICAT, " (", pie_data_descend$perc, "%)", sep="")

p2 = ggplot(pie_data_descend, aes(x= INDICAT, y=n, fill=INDICAT)) + 
    geom_bar(stat="identity") + 
    ggtitle(paste("Liver (n=", sum(pie_data_descend$n), ")", sep="" )) +
    coord_flip() +
    scale_fill_manual(values=col_indicat[pie_data_descend$INDICAT], drop=FALSE )+
    ylab("Number of samples") + xlab("Indication") + labs(caption = paste("", sep="")) +
    theme(legend.position="none")+
    scale_x_discrete(labels=addline_format(label_x) )+ 
    theme_bw()

#####

Design_i <- dplyr::filter(Design, grepl("ymph",BMFLOC, ignore.case = TRUE))
tally_i = Design_i %>% group_by( INDICAT) %>% tally()
tally_i$perc <- round((tally_i$n*100)/sum(tally_i$n), digits =1)
pie_data_descend <- tally_i %>% filter (n > n_min) %>% drop_na(INDICAT)
label_x = paste(pie_data_descend$INDICAT, " (", pie_data_descend$perc, "%)", sep="")


p3 = ggplot(pie_data_descend, aes(x= INDICAT, y=n, fill=INDICAT)) + 
    geom_bar(stat="identity") + 
    ggtitle(paste("Lymph node (n=", sum(pie_data_descend$n), ")", sep="" )) +
    coord_flip() +
    scale_fill_manual(values=col_indicat[pie_data_descend$INDICAT], drop=FALSE )+
    ylab("Number of samples") + xlab("Indication") + labs(caption = paste("Only indications with >", n_min, " samples are shown", sep="")) +
    theme(legend.position="none")+
    scale_x_discrete(labels=addline_format(label_x) )+ 
    theme_bw()


grid.arrange(p1, p2, p3, nrow = 1)


svg(paste(wd, "perTissue_HA.svg", sep="/"), width = 15, height = 4)
grid.arrange(p1, p2, p3, nrow = 1)
dev.off()

```

## Fisher test for CD8IMMPH and metastatic status

* Is there higher relative fraction of CD8-desert phenotypes in liver metastases across all tumor types?

```{r fisher mets, fig.height=5, fig.width=12, fig.align="center", message=FALSE, warning=FALSE}

Design <- Design %>% mutate( METSorNOT = ifelse(STATUS == "Primary", "Primary", "Metastasis"))

tmp = Design %>% filter(METSorNOT != "NA")
tmp2 = table(tmp [,c("CD8IMMPH","METSorNOT")])

kableExtra::kable(tmp2, caption = "Frequency table")

print("2x3 Fisher test")
fisher.test(tmp2, alternative = "two.sided")


a=fisher.test(tmp2[-1, ], alternative = "two.sided")
b=fisher.test(tmp2[-2, ], alternative = "two.sided")
c=fisher.test(tmp2[-3, ], alternative = "two.sided")

test_res= data.frame(comparison=c("Excluded_vs_Inflammed", "Desert_vs_Inflamed","Desert_vs_Excluded"), 
                     pvalue= c(a$p.value, b$p.value, c$p.value), 
                     odds_ratio = c(a$estimate, b$estimate, c$estimate))

DT::datatable(test_res, caption = "Odds ratios by 2x2") %>%
  formatRound(columns=c('pvalue', 'odds_ratio'), digits=5)

cat("\n")
cat("\n")
```

* There are some samples n= `r sum(is.na(Design$METSorNOT))`, that are NA in the STATUS because they do not have BFMLOC, so not possible to know whether they are primary or metastatic

# Tumor by location and immphen

```{r barplot 1, fig.height=5, fig.width=12, fig.align="center", message=FALSE, warning=FALSE}

full_table <- data.frame()

DT::datatable(Design_i %>% filter(STATUS == "Other Mets") %>% group_by(BMFLOC) %>% tally())

pheno_col <- c(
  INFLAMED = "#d95f02",
  EXCLUDED = "#1b9e77",
  DESERT = "#7570b3"
)

Design_i <- dplyr::filter(Design, grepl("UBC",INDICAT, ignore.case = TRUE))

tally_i = Design_i %>% group_by(CD8IMMPH, STATUS) %>% tally()
tally_i$STATUS <- factor(tally_i$STATUS, levels = c("Primary","Liver Mets","Lung Mets" ,"LN Mets" ,"Kidney Mets" ,"Colon Mets","HN Mets","Breast Mets","Ovarian Mets","Pancreas Mets","Other Mets") )

tally_i <- na.omit(tally_i)
to_other <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency <=4) %>% filter(grepl("Met",STATUS))
tally_i[tally_i$STATUS%in%to_other$STATUS,"STATUS"]="Other Mets"
to_keep <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency >4)
tally_i <- tally_i %>% filter(STATUS %in% to_keep$STATUS)

label_x = tally_i %>% group_by(STATUS) %>% add_count(STATUS, wt = n) %>% dplyr::select(STATUS, nn) %>% distinct()
label_x = label_x[ match( levels(tally_i$STATUS), label_x$STATUS), ]
label_x <- na.omit(label_x)
label_x = paste(label_x$STATUS, " (", label_x$nn, ")", sep="")
#label_x = label_x[c(6,2,4,3,1,5)]


p1 = ggplot(tally_i, aes(fill= CD8IMMPH, y=n, x=STATUS)) + 
    geom_bar(position="fill", stat="identity") + 
    theme(legend.position="top") + 
    ggtitle("UBC") +
    labs(fill = "")+
    scale_fill_manual(values=pheno_col) +
    ylab("Frequency (%)") + xlab("") +
    labs(caption = paste("Plot only lesions >5 samples, total samples=",sum(tally_i$n), sep=""))+
    scale_x_discrete(labels=addline_format(label_x) ) +
    theme_bw()


full_table <- rbind( full_table, 
                     tally_i %>% mutate(TumorExcision= "BLADDER"))

# st=paste("total samples=",sum(tally_i$n) ,". Only mets from liver, lung, LN, kidney, colon, HN, breast, ovarian, pancreas are separated. All other are collapsed in other mets.", sep="")



#####

Design_i <- dplyr::filter(Design, grepl("NSCLC",INDICAT, ignore.case = TRUE))
Design_i$STATUS[  which( grepl("LUNG", Design_i$BMFLOC))] = "Primary"
Design_i$STATUS[  which( grepl("lung", Design_i$BMFLOC))] = "Primary"

tally_i = Design_i %>% group_by(CD8IMMPH, STATUS) %>% tally()
tally_i$STATUS <- factor(tally_i$STATUS, levels = c("Primary","Liver Mets","Lung Mets" ,"LN Mets" ,"Kidney Mets" ,"Colon Mets","HN Mets","Breast Mets","Ovarian Mets","Pancreas Mets","Other Mets") )

tally_i <- na.omit(tally_i)
to_other <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency <=4) %>% filter(grepl("Met",STATUS))
tally_i[tally_i$STATUS%in%to_other$STATUS,"STATUS"]="Other Mets"
to_keep <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency >4)
tally_i <- tally_i %>% filter(STATUS %in% to_keep$STATUS)

label_x = tally_i %>% group_by(STATUS) %>% add_count(STATUS, wt = n) %>% dplyr::select(STATUS, nn) %>% distinct()
# primary, liver, lung, LN, kidney, other
#label_x = label_x[c(5,2,3,1,4)]
label_x = label_x[ match( levels(tally_i$STATUS), label_x$STATUS), ]
label_x <- na.omit(label_x)
label_x = paste(label_x$STATUS, " (", label_x$nn, ")", sep="")

p2 = ggplot(tally_i, aes(fill= CD8IMMPH, y=n, x=STATUS)) + 
    geom_bar(position="fill", stat="identity") + 
    theme(legend.position="top") + 
    ggtitle("NSCLC") +
    labs(fill = "")+
    scale_fill_manual(values=pheno_col) +
    ylab("Frequency (%)") + xlab("") +
    labs(caption = paste("Plot only lesions >=5 samples, total samples=",sum(tally_i$n) , sep="")) +
    scale_x_discrete(labels=addline_format(label_x) ) +
    theme_bw()  

full_table <- rbind( full_table, 
                     tally_i %>% mutate(TumorExcision= "LUNG"))

#####

Design_i <- dplyr::filter(Design, grepl("CRC",INDICAT, ignore.case = TRUE))
Design_i$STATUS[  which( grepl("COLON", Design_i$BMFLOC))] = "Primary"

tally_i = Design_i %>% group_by(CD8IMMPH, STATUS) %>% tally()
tally_i$STATUS <- factor(tally_i$STATUS, levels = c("Primary","Liver Mets","Lung Mets" ,"LN Mets" ,"Kidney Mets" ,"Colon Mets","HN Mets","Breast Mets","Ovarian Mets","Pancreas Mets","Other Mets") )

tally_i <- na.omit(tally_i)
to_other <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency <=4) %>% filter(grepl("Met",STATUS))
tally_i[tally_i$STATUS%in%to_other$STATUS,"STATUS"]="Other Mets"
to_keep <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency >4)
tally_i <- tally_i %>% filter(STATUS %in% to_keep$STATUS)

label_x = tally_i %>% group_by(STATUS) %>% add_count(STATUS, wt = n) %>% dplyr::select(STATUS, nn) %>% distinct()
# primary, liver, lung, LN, kidney, other
#label_x = label_x[c(5,2,3,1,4)]
label_x = label_x[ match( levels(tally_i$STATUS), label_x$STATUS), ]
label_x <- na.omit(label_x)
label_x = paste(label_x$STATUS, " (", label_x$nn, ")", sep="")

p3 = ggplot(tally_i, aes(fill= CD8IMMPH, y=n, x=STATUS)) + 
    geom_bar(position="fill", stat="identity") + 
    theme(legend.position="top") + 
    ggtitle("CRC") +
    labs(fill = "")+
    scale_fill_manual(values=pheno_col) +
    ylab("Frequency (%)") + xlab("") +
    labs(caption = paste("Plot only lesions >5 samples, total samples=",sum(tally_i$n) , sep="")) +
    scale_x_discrete(labels=addline_format(label_x) ) + 
    theme_bw()

full_table <- rbind( full_table, 
                     tally_i %>% mutate(TumorExcision= "COLORECTAL"))

#####

grid.arrange(p2, p3, p1, nrow = 1)



svg(paste(wd, "perSTATUS_HA.svg", sep="/"), width = 15, height = 6)
grid.arrange(p2, p3, p1, nrow = 1)
dev.off()


#####

indic_i = "RCC"

Design_i <- dplyr::filter(Design, grepl(indic_i,INDICAT, ignore.case = TRUE))
Design_i$STATUS[  which( grepl(indic_i, Design_i$BMFLOC))] = "Primary"

tally_i = Design_i %>% group_by(CD8IMMPH, STATUS) %>% tally()
tally_i$STATUS <- factor(tally_i$STATUS, levels = c("Primary","Liver Mets","Lung Mets" ,"LN Mets" ,"Kidney Mets" ,"Colon Mets","HN Mets","Breast Mets","Ovarian Mets","Pancreas Mets","Other Mets") )

tally_i <- na.omit(tally_i)
to_other <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency <=4) %>% filter(grepl("Met",STATUS))
tally_i[tally_i$STATUS%in%to_other$STATUS,"STATUS"]="Other Mets"

to_keep <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency >4)
tally_i <- tally_i %>% filter(STATUS %in% to_keep$STATUS)

label_x = tally_i %>% group_by(STATUS) %>% add_count(STATUS, wt = n) %>% dplyr::select(STATUS, nn) %>% distinct()
# primary, liver, lung, LN, kidney, other
#label_x = label_x[c(5,2,3,1,4)]
label_x = label_x[ match( levels(tally_i$STATUS), label_x$STATUS), ]
label_x <- na.omit(label_x)
label_x = paste(label_x$STATUS, " (", label_x$nn, ")", sep="")

p4 = ggplot(tally_i, aes(fill= CD8IMMPH, y=n, x=STATUS)) + 
    geom_bar(position="fill", stat="identity") + 
    theme(legend.position="top") + 
    ggtitle(paste(indic_i, sep=" ")) +
    labs(fill = "")+
    scale_fill_manual(values=pheno_col) +
    ylab("Frequency (%)") + xlab("") +
    labs(caption = paste("Plot only lesions >5 samples, total samples=",sum(tally_i$n) , sep="")) +
    scale_x_discrete(labels=addline_format(label_x) ) + 
    theme_bw()


full_table <- rbind( full_table, 
                     tally_i %>% mutate(TumorExcision= indic_i))

#####

#####

indic_i = "HNC"

Design_i <- dplyr::filter(Design, grepl(indic_i,INDICAT, ignore.case = TRUE))
Design_i$STATUS[  which( grepl(indic_i, Design_i$BMFLOC))] = "Primary"

tally_i = Design_i %>% group_by(CD8IMMPH, STATUS) %>% tally()
tally_i$STATUS <- factor(tally_i$STATUS, levels = c("Primary","Liver Mets","Lung Mets" ,"LN Mets" ,"Kidney Mets" ,"Colon Mets","HN Mets","Breast Mets","Ovarian Mets","Pancreas Mets","Other Mets") )

tally_i <- na.omit(tally_i)
to_other <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency <=4) %>% filter(grepl("Met",STATUS))
tally_i[tally_i$STATUS%in%to_other$STATUS,"STATUS"]="Other Mets"

to_keep <- tally_i %>% group_by(STATUS) %>% summarise(Frequency = sum(n)) %>% filter(Frequency >4)
tally_i <- tally_i %>% filter(STATUS %in% to_keep$STATUS)

label_x = tally_i %>% group_by(STATUS) %>% add_count(STATUS, wt = n) %>% dplyr::select(STATUS, nn) %>% distinct()
# primary, liver, lung, LN, kidney, other
#label_x = label_x[c(5,2,3,1,4)]
label_x = label_x[ match( levels(tally_i$STATUS), label_x$STATUS), ]
label_x <- na.omit(label_x)
label_x = paste(label_x$STATUS, " (", label_x$nn, ")", sep="")

p5 = ggplot(tally_i, aes(fill= CD8IMMPH, y=n, x=STATUS)) + 
    geom_bar(position="fill", stat="identity") + 
    theme(legend.position="top") + 
    ggtitle(paste(indic_i, sep=" ")) +
    labs(fill = "")+
    scale_fill_manual(values=pheno_col) +
    ylab("Frequency (%)") + xlab("") +
    labs(caption = paste("Plot only lesions >5 samples, total samples=",sum(tally_i$n) , sep="")) +
    scale_x_discrete(labels=addline_format(label_x) ) + 
    theme_bw()

full_table <- rbind( full_table, 
                     tally_i %>% mutate(TumorExcision= indic_i))



DT::datatable(full_table %>% pivot_wider(names_from = CD8IMMPH, values_from = n) %>% relocate(TumorExcision), options=list(buttons=c('excel')))

#####
```


```{r , fig.height=5, fig.width=20, fig.align="center", message=FALSE, warning=FALSE}


svg(paste(wd, "perSTATUS_onlyKIDNEY_HN_HA.svg", sep="/"), width = 10, height = 5)
grid.arrange(p4, p5, nrow = 1)
dev.off()

grid.arrange(p2, p3, p1, p4,p5, nrow = 1)

svg(paste(wd, "perSTATUS_All_HA.svg", sep="/"), width = 25, height = 5)
grid.arrange(p2, p3, p1, p4,p5, nrow = 1)
dev.off()

```



# Tumor by studyID and indication 

```{r barplot 2, fig.height=5, fig.width=12, fig.align="center", message=FALSE, warning=FALSE}

Design_i <- Design

# solve that primary vs all mets category is not present 

tally_i = Design_i %>% group_by(STUDYID, INDICAT) %>% tally()
pie_data_descend <- tally_i %>% filter (n > 5) %>% drop_na(INDICAT)

label_x = tally_i %>% group_by(STUDYID) %>% add_count(STUDYID, wt = n) %>% dplyr::select(STUDYID, nn) %>% distinct()
label_x = paste(label_x$STUDYID, " (", label_x$nn, ")", sep="")

p1= ggplot(pie_data_descend, aes(fill= INDICAT, y=n, x=STUDYID)) + 
    geom_bar(position="fill", stat="identity") + 
    ggtitle("Indications per study") +
    scale_fill_manual(values=colorRampPalette(brewer.pal(8,"Dark2"))(length(unique(pie_data_descend$INDICAT))) ) +
    scale_x_discrete(labels=addline_format(label_x) ) +
    ylab("Frequency (%)") + xlab("")+ 
    theme_bw()

p1

svg(paste(wd, "Studies_per_indicat_HA.svg", sep="/"), width = 15, height = 6)
p1
dev.off()

```
