---
title: "Feature distribution and model export"
author: "Iakov Davydov"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  data_collection: "UUID"
  output_collection: "UUID"
  output_sharing_collection: "UUID"
  .arv_save:
    value:
      collection: "UUID"
      filename: "{basename(html_filename)}"
---

Goals:
1. For features with non-zero coefficients we would like to save their
  distribution for basic QC of input data.
2. We would like to export the model w/o the mlr wrapper for easier reuse.

```{r libs, message=FALSE, warning=FALSE}
library(glmnet)
library(tidyverse)
library(mlr3verse)
```

```{r funcs}
source("../src/eset_transformation.R")
source("../src/mlr_helpers.R")
source("../src/pv.R")
source("../src/density.R")
```


```{r "random seed"}
set.seed(42)
```

Load ExpressionSet object.
```{r "load eset"}
eset_train <- aws.s3::s3readRDS("eset_train.Rds", params$output_collection)
eset_validation <- aws.s3::s3readRDS("eset_validation.Rds", params$output_collection)
eset_test <- aws.s3::s3readRDS("eset_test.Rds", params$output_collection)
eset_fake <- aws.s3::s3readRDS("eset_fake_counts.Rds", params$output_sharing_collection)
```

Load TCGA
```{r}
tcga_eset <- aws.s3::s3readRDS(
  "tcga_eset_batch_corrected.Rds",
  bucket = params$output_collection
)
```


Combine ExpressionSet objects
```{r "combine eset"}
eset_trval <- BiocGenerics::combine(
  eset_train,
  eset_validation
)

stopifnot(nrow(eset_train) == nrow(eset_validation))
stopifnot(nrow(eset_train) == nrow(eset_trval))
stopifnot(ncol(eset_train) + ncol(eset_validation) == ncol(eset_trval))
```


Re-create tibbles for training.
```{r "create training and test df"}
dataset <- create_modeling_df(eset_trval)
dataset_test <- create_modeling_df(eset_test)
Biobase::exprs(eset_fake) <- DESeq2::vst(
  Biobase::exprs(eset_fake)
)
dataset_fake <- create_modeling_df(eset_fake)
```

Model
```{r}
model <- aws.s3::s3readRDS(
  object = "full_model.Rds",
  bucket = params$output_collection
)
```

Features
```{r}
features <- extr_glm_coef(model) %>%
  filter(feature != "(Intercept)") %>%
  pull(feature) %>%
  unique()
```


```{r fig.width=10, fig.height=8}
feature_values <- dataset %>%
  select(all_of(features))

feature_values %>%
  pivot_longer(everything()) %>%
  ggplot() +
  aes(value) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density(bw = "SJ", col = "#ff202080") +
  facet_wrap(~name, scales = "free") +
  theme_bw() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

```{r}
feature_values_sim <- feature_values %>%
  map(simulate_from_density, 1e5) %>%
  as_tibble()
```


Check that density does match
```{r}
bind_rows(
  observed = feature_values,
  sim = feature_values_sim,
  .id = "type"
) %>%
  .[, 1:10] %>%
  pivot_longer(-type) %>%
  ggplot() +
  aes(value, color = type, linetype = type) +
  geom_density() +
  facet_wrap(~ name, scales = "free_x") +
  theme_bw()
```

Check if TCGA data matches the training dataset distribution
```{r}
model_genes <- model$selected_features()

stopifnot(all(rownames(tcga_eset) == rownames(eset_train)))
stopifnot(nrow(tcga_eset) == nrow(eset_train))
Biobase::fData(tcga_eset)$symbol <- Biobase::fData(eset_train)$symbol
tcga_eset$CD8IMMPH <- NA

tcga_df <- create_modeling_df(tcga_eset)
```


```{r}
reference <- feature_values_sim %>%
  map(ecdf)
ecdf_pvals(dataset, reference) %>%
  summary()
ecdf_pvals(dataset_fake, reference) %>%
  summary()
ecdf_pvals(dataset_test, reference) %>%
  summary()
ecdf_pvals(tcga_df, reference) %>%
  summary()
```

```{r "save data into Rds object"}
aws.s3::s3saveRDS(
  feature_values_sim,
  "feature_distribution.Rds",
  bucket = params$output_sharing_collection
)
```

```{r "export the glmnet model"}
stopifnot(colnames(model$state$data_prototype)[1] == "IHC_Status")
m <- list(
  cnames = tail(colnames(model$state$data_prototype), -1),
  model = model$state$model
)

aws.s3::s3saveRDS(
  m,
  "full_model_glmnet.Rds",
  bucket = params$output_sharing_collection
)
```

```{r, results="asis", echo=FALSE}
reporting::print_reproducibility_section(params)
```
