---
title: "Survival plots based on immune phenotype and model prediction"
author: "Iakov Davydov"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  data_collection: "UUID"
  output_collection: "UUID"
  .arv_save:
    value:
      collection: "UUID"
      filename: "{basename(html_filename)}"
---

```{r libs, message=FALSE, warning=FALSE}
library(survival)
library(survminer)
library(glmnet)
library(tidyverse)
library(mlr3verse)
```

```{r funcs}
source("../src/eset_transformation.R")
source("../src/mlr_helpers.R")
source("../src/utils.R")
```

```{r}
theme_set(theme_bw())
```

```{r "random seed"}
set.seed(42)
```

Load ExpressionSet object.
```{r "load eset"}
eset <- aws.s3::s3readRDS("eset_batch_corrected.Rds", params$output_collection)
eset_tpm <- aws.s3::s3readRDS("tpm_eset.Rds", params$output_collection)
```

Compute T eff signature
```{r "compute t eff signature"}
t_eff <- c(
  "CD3E",
  "CD8A",
  "CXCL10",
  "CXCL9",
  "EOMES",
  "GZMA",
  "GZMB",
  "IFNG",
  "KLRD1",
  "NKG7",
  "PRF1",
  "TBX21"
)
t_eff_geneids <- Biobase::fData(eset_tpm) %>%
  rownames_to_column("GeneID") %>%
  filter(symbol %in% t_eff) %>%
  {
    set_names(.$GeneID, .$symbol)
  }

sign <- BioQC::GmtList(list(t_eff = list(name = "t_eff", desc = "t_eff", genes = t_eff)))

sig_scores <- BioQC::wmwTest(eset_symbol_matrix(eset_tpm),
  sign,
  valType = "r"
)




sig_scores <- tibble(A_SampleID = names(sig_scores), t_eff = sig_scores)
sig_scores$CD8A <- Biobase::exprs(eset_tpm[t_eff_geneids["CD8A"], ])[, sig_scores$A_SampleID]
sig_scores$t_eff_tpm <- eset_tpm[t_eff_geneids, ] %>%
  Biobase::exprs() %>%
  colSums() %>%
  {
    .[sig_scores$A_SampleID]
  }
```


Load models
```{r "load model"}
model <- aws.s3::s3readRDS(
  object = "full_model.Rds",
  bucket = params$output_collection
)
```

```{r "get data source annotation"}
full_sample_anno <- aws.s3::s3read_using(
  read_tsv,
  show_col_types = FALSE,
  object = "full_samples_anno.tsv",
  bucket = params$data_collection
)
```

Create tibbles for validation
```{r "create validation df"}
dataset <- create_modeling_df(eset)
```

```{r predict}
pred <- model$predict_newdata(dataset)
```

```{r "prepare performance dfs"}
pred_df <- as.data.table(pred)
pred_df$A_SampleID <- colnames(eset)
pred_df <- left_join(
  pred_df,
  select(full_sample_anno, A_SampleID, source = SOURCE, study = STUDYID),
  by = "A_SampleID"
)
pred_df$truth <- fct_relevel(pred_df$truth, names(pheno_col))
pred_df$response <- fct_relevel(pred_df$response, names(pheno_col))
pred_df <- left_join(
  pred_df,
  sig_scores,
  by = "A_SampleID"
)
pred_df$logit.inflamed <- with(pred_df, log(prob.inflamed / (1 - prob.inflamed)))
```

# Survival plots {.tabset .tabset-pills}
```{r "CD8 imm phe OS dataset"}
os_data <- filter(full_sample_anno, !is.na(OS) & !is.na(ARM)) %>%
  left_join(pred_df, by = "A_SampleID") %>%
  rename(prediction = response)

# Censoring Status Value(1=cens,0=evt)
# https://drive.google.com/file/d/11oG4_rXtEzBkJmWC9x009Z_JIIga1uMc/view
os_data$event <- if_else(os_data$CNSR_OS == 1, 0, 1)

os_data$atezo <- if_else(str_detect(os_data$ARM, fixed("MPDL3280A")), TRUE, FALSE)
```

```{r}
study_names <- c(
  "GO29293" = "IMvigor 210",
  "GO28915" = "OAK",
  "GO29294" = "IMvigor 211"
)
```


```{r "survplot_study"}
survplot_dataset <- function(os_data, fit, atezo) {
  pal <- unname(pheno_col[levels(os_data$truth)])
  ggsurvplot(fit,
    data = os_data, risk.table = TRUE, risk.table.height = 0.25,
    palette = pal,
    conf.int = TRUE,
    title = if (!is.na(atezo)) str_glue("atezo={atezo}") else ""
  )
}

survplot_studies <- function(os_data) {
  frm1 <- Surv(OS, event) ~ truth
  frm2 <- Surv(OS, event) ~ prediction
  d <- os_data %>%
    group_by(study, atezo) %>%
    nest()
  d$fit_truth <- surv_fit(frm1, d$data)
  d$fit_prediction <- surv_fit(frm2, d$data)
  d <- d %>%
    pivot_longer(starts_with("fit_"), names_prefix = "fit_", names_to = "model", values_to = "fit")

  d <- d %>%
    group_by(study, model) %>%
    mutate(n_atezo = length(unique(atezo))) %>%
    ungroup()
  d <- d %>%
    mutate(atezo = if_else(n_atezo == 1, NA, atezo))
  d$plt <- pmap(d, function(data, fit, atezo, ...) survplot_dataset(data, fit, atezo))
  d
}

plots <- survplot_studies(os_data)
```

```{r}
oak_atezo_data <- os_data %>%
  filter(
    study == names(study_names)[study_names == "OAK"],
    atezo == TRUE
  ) 
oak_atezo_fit <-
  surv_fit(Surv(OS, event) ~ prediction, oak_atezo_data)

{
  svg(
    here::here("figures/fig6c2_oak_survival_plot.svg"),
    width = 7.5,
    height = 6
  )
  survplot_dataset(oak_atezo_data, oak_atezo_fit, NA) %>%
    print(newpage = FALSE)
  dev.off()
}
```

```{r fig}
pal <- unname(pheno_col[levels(oak_atezo_data$truth)])
thm <-  theme(
  text = element_text(size = 16),
  legend.key.height = unit(3, "mm"),
  panel.background = element_rect(fill = "transparent"),
  strip.background = element_rect(fill = "transparent"),
  plot.background = element_rect(fill = "transparent", color = NA),
  legend.background = element_rect(fill = "transparent"),
  legend.key = element_rect(fill = "transparent"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()
)
oak_atezo_fit_tmp <- oak_atezo_fit
names(oak_atezo_fit_tmp$strata) <- 
  str_replace(names(oak_atezo_fit_tmp$strata), ".*=", "")
{
  svg(
    here::here("figures/fig6c2_oak_survival_plot_poster.svg"),
    width = 4.5,
    height = 4.5
  )
  ggsurvplot(oak_atezo_fit_tmp,
    data = oak_atezo_data, risk.table = TRUE, risk.table.height = .35,
    palette = pal,
    conf.int = TRUE,
    ggtheme = thm
  ) %>%
    print(newpage = FALSE)
  dev.off()
}
```


```{r "os plots", fig.width=7, fig.height=6, warning=FALSE, results="asis"}
for (s in unique(plots$study)) {
  cat(str_glue("## {s} ({study_names[s]}) {{.tabset .tabset-pills}}\n\n", trim = FALSE))
  cat(str_glue("### truth {{.tabset .tabset-pills}}\n\n", trim = FALSE))
  d <- filter(plots, study == s & model == "truth")
  for (a in d$atezo) {
    if (!is.na(a)) {
      dd <- filter(d, atezo == a)
      cat(str_glue("#### atezo={a} {{.tabset .tabset-pills}}\n\n", trim = FALSE))
    } else {
      dd <- d
    }
    dd$plt %>% walk(print)
    cat("\n\n")
  }

  cat(str_glue("### prediction {{.tabset .tabset-pills}}\n\n", trim = FALSE))
  d <- filter(plots, study == s & model == "prediction")
  for (a in d$atezo) {
    if (!is.na(a)) {
      dd <- filter(d, atezo == a)
      cat(str_glue("#### atezo={a} {{.tabset .tabset-pills}}\n\n", trim = FALSE))
    } else {
      dd <- d
    }
    dd$plt %>% walk(print)
    cat("\n\n")
  }
}
```


## OAK plots {.tabset .tabset-pills}

```{r "prepare OS data OAK"}
os_data_oak <- filter(os_data, study == names(study_names)[study_names == "OAK"])
os_data_oak$t_eff_high <- os_data_oak$t_eff > median(os_data_oak$t_eff)
os_data_oak$t_eff_sum_high <- os_data_oak$t_eff_tpm > median(os_data_oak$t_eff_tpm)
os_data_oak$cd8a_high <- os_data_oak$CD8A > median(os_data_oak$CD8A)
os_data_oak$predicted_inflamed <- os_data_oak$prediction == "inflamed"
os_data_oak$truth_inflamed <- os_data_oak$truth == "inflamed"
os_data_oak$predicted_inflamed_subset <- if_else(
  is.na(os_data_oak$truth_inflamed),
  NA,
  os_data_oak$predicted_inflamed
)

plot_os_oak <- function(data, response, include_atezo) {
  data <- data[data$atezo == include_atezo, ]
  frm <- as.formula(str_c("Surv(OS, event) ~ ", response))
  fit <- surv_fit(frm, data = data)
  ggsurvplot(fit,
    title = str_glue("atezo={include_atezo}"),
    data = data, risk.table = TRUE, risk.table.height = 0.25,
    conf.int = TRUE
  )
}
```


### Predicted inflamed, atezo arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "predicted_inflamed", TRUE)
```

### Predicted inflamed, docetaxel arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "predicted_inflamed", FALSE)
```

### Predicted inflamed (subset with truth), atezo arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "predicted_inflamed_subset", TRUE)
```

### Predicted inflamed (subset with truth), docetaxel arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "predicted_inflamed_subset", FALSE)
```

### Truth inflamed, atezo arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "truth_inflamed", TRUE)
```

### Truth inflamed, docetaxel arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "truth_inflamed", FALSE)
```

### T effector high, atezo arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "t_eff_high", TRUE)
```

### T effector high, docetaxel arm
```{r fig.width=7, fig.height=6, warning=FALSE, results="asis"}
plot_os_oak(os_data_oak, "t_eff_high", FALSE)
```


### Forest plots
```{r fig.width=8, fig.height=3}
fit_to_df <- function(f, d) {
  s <- summary(f)
  tibble(
    HR = s$conf.int[, "exp(coef)"],
    HR_lower = s$conf.int[, "lower .95"],
    HR_upper = s$conf.int[, "upper .95"],
    n = s$n,
    n_events = s$nevent
  )
}


coxph_models <- crossing(
  arm = c("atezo", "docetaxel"),
  response = c("predicted_inflamed", "predicted_inflamed_subset", "truth_inflamed", "t_eff_high")
) %>%
  mutate(response_label = case_when(
    response == "predicted_inflamed" ~ "Predicted: inflamed",
    response == "predicted_inflamed_subset" ~ "Predicted: inflamed (subset)",
    response == "truth_inflamed" ~ "Truth: inflamed",
    response == "t_eff_high" ~ "T effector high")) %>%
  rowwise() %>%
  mutate(
    data = list(os_data_oak[os_data_oak$atezo == (arm == "atezo"), ]),
    fit = list(
      survival::coxph(as.formula(str_c("Surv(OS, event) ~ ", response)),
        data = data
      )
    ),
    fit_summary = list(fit_to_df(fit))
  ) %>%
  unnest(fit_summary)

m <- with(coxph_models, meta::metagen(log(HR),
  lower = log(HR_lower), upper = log(HR_upper),
  sm = "HR",
  studlab = str_glue("ARM {arm} (N={n}), {response_label}"),
  data = coxph_models
))


meta::forest(m,
  xlim = c(0.5, 2), pooled.totals = TRUE, lab.e = "events",
  overall = FALSE,
  leftlabs = c(""), leftcols = "studlab",
  hetstat = FALSE,
)
```

```{r fig.width=8, fig.height=3}
# only predicted

m <- coxph_models %>%
  filter(response %in% c("predicted_inflamed")) %>%
  with(., meta::metagen(log(HR),
  lower = log(HR_lower), upper = log(HR_upper),
  sm = "HR",
  studlab = str_glue("ARM {arm} (N={n}), {response_label}"),
  data = .
))

meta::forest(m,
  xlim = c(0.5, 2), pooled.totals = TRUE, lab.e = "events",
  overall = FALSE,
  leftlabs = c(""), leftcols = "studlab",
  hetstat = FALSE,
)
{
  svg(
    here::here("figures/fig6c1_oak_survival_forest.svg"),
    width = 8,
    height = 2.1
  )
  meta::forest(m,
    xlim = c(0.5, 2), pooled.totals = TRUE, lab.e = "events",
    overall = FALSE,
    leftlabs = c(""), leftcols = "studlab",
    hetstat = FALSE,
  )
  dev.off()
}
```

# Correlation between inflamed probability and T effector signature scores
```{r "correlation between inflamed probability and T eff"}
ggplot(pred_df) +
  aes(prob.inflamed, t_eff, col = truth) +
  geom_point() +
  scale_color_manual(values = pheno_col) +
  ggtitle("Full dataset") +
  guides(color = guide_legend(title = "CD8 immune phenotype"))

pred_df %>%
  filter(study == "GO28915") %>%
  ggplot() +
  aes(prob.inflamed, t_eff, col = truth) +
  geom_point() +
  scale_color_manual(values = pheno_col) +
  ggtitle("OAK") +
  guides(color = guide_legend(title = "CD8 immune phenotype"))
```

```{r "boxplot of T effector vs prediction"}
pred_df %>%
  ggplot(aes(t_eff, response, fill = response)) +
  geom_boxplot()
```

TPMs below are log2 transformed (with pseudocount).
```{r "correlate T effector to inflamed probability"}
pred_df %>%
  select(logit.inflamed, CD8A, t_eff, t_eff_tpm) %>%
  mutate(CD8A = log2(CD8A + 1), t_eff_tpm = log2(t_eff_tpm + 1)) %>%
  GGally::ggpairs(
    progress = FALSE,
    upper = list(continuous = "density"),
    lower = list(continuous = GGally::wrap("points", alpha = 0.1))
  )
```

```{r, results="asis", echo=FALSE}
reporting::print_reproducibility_section(params)
```
