---
title: "Apply model to TCGA"
author: "Iakov Davydov"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  data_collection: "UUID"
  output_collection: "UUID"
  .arv_save:
    value:
      collection: "UUID"
      filename: "{basename(html_filename)}"
---

```{r libs, message=FALSE, warning=FALSE}
library(tidyverse)
library(mlr3verse)
library(glmnet)
library(RTCGA.clinical)
library(Biobase)
```

```{r ggtheme}
theme_set(theme_bw())
```


```{r}
source("../src/eset_transformation.R")
source("../src/utils.R")
```

```{r}
theme_set(theme_bw())
```

```{r "random seed"}
set.seed(42)
```

Load models
```{r "load model"}
model <- aws.s3::s3readRDS(
  object = "full_model.Rds",
  bucket = params$output_collection
)
```

```{r "load tcga and roche data"}
tcga_eset <- aws.s3::s3readRDS(
  "tcga_eset_batch_corrected.Rds",
  bucket = params$output_collection
)
eset_train <- aws.s3::s3readRDS(
  "eset_train.Rds",
  params$output_collection
)
```


Check feature distribution
```{r "feature correspondance"}
model_genes <- model$selected_features()

stopifnot(all(rownames(tcga_eset) == rownames(eset_train)))
stopifnot(nrow(tcga_eset) == nrow(eset_train))
fData(tcga_eset)$symbol <- fData(eset_train)$symbol
tcga_eset$CD8IMMPH <- NA

tcga_df <- create_modeling_df(tcga_eset)
```

```{r}
eset_rows_to_use <- fData(eset_train)$symbol %in% model_genes
stopifnot(eset_rows_to_use == fData(tcga_eset)$symbol %in% model_genes)
eset_m <- exprs(eset_train)[eset_rows_to_use, ]
tcga_m <- exprs(tcga_eset)[eset_rows_to_use, ]

stopifnot(nrow(eset_m) == nrow(tcga_m))
```

Heatmap
```{r "heatmap", fig.width=8, fig.height=8}
m <- cbind(eset_m, tcga_m)
rownames(m) <- fData(eset_train)$symbol[eset_rows_to_use]

cannot <- ComplexHeatmap::columnAnnotation(
  dataset = c(rep("train", ncol(eset_m)), rep("TCGA", ncol(tcga_m))),
  IHC_Status = c(str_to_lower(eset_train$CD8IMMPH), rep(NA, ncol(tcga_m))),
  col = list(IHC_Status = pheno_col),
  study = c(rep(NA, ncol(eset_m)), tcga_eset$study)
)

ComplexHeatmap::Heatmap(m,
  top_annotation = cannot,
  show_column_names = FALSE,
  show_row_names = FALSE,
  cluster_columns = FALSE
)
```

# Predicting for TCGA
```{r}
pred <- model$predict_newdata(select(tcga_df, -IHC_Status))
tcga_eset$CD8IMMPH_pred <- pred$response
```

# Prediction distribution per indication
```{r}
tcga_eset_summary <- pData(tcga_eset) %>%
  select(study, CD8IMMPH_pred) %>%
  group_by(study) %>%
  mutate(prop_inflamed = sum(CD8IMMPH_pred == "inflamed") / n(), count = n()) %>%
  ungroup() %>%
  mutate(study = fct_reorder(study, -prop_inflamed))

p <- tcga_eset_summary %>%
  ggplot() +
  aes(study, fill = CD8IMMPH_pred) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = pheno_col, name = "predicted phenotype") +
  geom_text(aes(label = count, x = study, fill = NULL),
    data = distinct(tcga_eset_summary, study, count),
    y = 1.01, angle = 90, hjust = 0
  ) +
  ylim(0, 1.09) +
  ylab("proportion")
p

ggsave(here::here("figures/tcga_predicted.svg"))

```

```{r}
p$layers[[2]]$aes_params$size <- 2
p$layers[[2]]$aes_params$y <- 1.02 
p$coordinate
p <- p +
  theme(
    text = element_text(size = 6),
    legend.key.height = unit(3, "mm"),
    #panel.background = element_rect(fill = "transparent"),
    strip.background = element_rect(fill = "transparent"),
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent"),
    legend.key = element_rect(fill = "transparent"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.box.margin=margin(-11,-11,-11,-11)
  ) +
  ylim(0, 1.25)
ggsave(
  here::here("figures/tcga_predicted_poster.png"),
  p,
  width = 7, height = 5, units = "cm", dpi = 300, scale = 1,
  bg = "transparent"
)
```

## Table
```{r}
tcga_eset_summary %>%
  group_by(study) %>%
  count(CD8IMMPH_pred) %>%
  mutate(prc = round(100 * n / sum(n), 1)) %>%
  pivot_wider(
    names_from = CD8IMMPH_pred,
    values_from = c(n, prc),
    values_fill = 0
  ) %>%
  arrange(-prc_inflamed, -prc_excluded, -prc_inflamed) %>%
  gt::gt()
```


# Survival analysis
```{r}
# https://github.com/RTCGA/RTCGA/issues/97#data-preparation
TCGA_surv <- survivalTCGA(
  ACC.clinical,
  BLCA.clinical,
  BRCA.clinical,
  CESC.clinical,
  CHOL.clinical,
  COAD.clinical,
  COADREAD.clinical,
  DLBC.clinical,
  ESCA.clinical,
  FPPP.clinical,
  GBM.clinical,
  GBMLGG.clinical,
  HNSC.clinical,
  KICH.clinical,
  KIPAN.clinical,
  KIRC.clinical,
  KIRP.clinical,
  LAML.clinical,
  LGG.clinical,
  LIHC.clinical,
  LUAD.clinical,
  LUSC.clinical,
  MESO.clinical,
  OV.clinical,
  PAAD.clinical,
  PCPG.clinical,
  PRAD.clinical,
  READ.clinical,
  SARC.clinical,
  SKCM.clinical,
  STAD.clinical,
  STES.clinical,
  TGCT.clinical,
  THCA.clinical,
  THYM.clinical,
  UCEC.clinical,
  UCS.clinical,
  UVM.clinical
)

tcga_subset <- tcga_eset[
  ,
  !is.na(tcga_eset$tcga.xml_bcr_patient_barcode) &
    !is.na(tcga_eset$tcga.cgc_sample_sample_type) &
    tcga_eset$tcga.xml_bcr_patient_barcode %in% TCGA_surv$bcr_patient_barcode &
    tcga_eset$tcga.cgc_sample_sample_type == "Primary Tumor"
]
tcga_subset_pred <- pData(tcga_subset) %>%
  select(tcga.xml_bcr_patient_barcode, CD8IMMPH_pred, cohort = study)

TCGA_surv_imm_phe <- TCGA_surv %>%
  inner_join(tcga_subset_pred, by = c("bcr_patient_barcode" = "tcga.xml_bcr_patient_barcode")) %>%
  rename(time = times, event = patient.vital_status)

table(TCGA_surv_imm_phe$cohort, useNA = "always")
```

```{r "survival plots", message=FALSE, warning=FALSE}
survplots <- unique(TCGA_surv_imm_phe$cohort) %>%
  set_names() %>%
  map(
    ~ RTCGA::kmTCGA(
      filter(TCGA_surv_imm_phe, cohort == .x),
      times = "time",
      status = "event",
      explanatory.names = c("CD8IMMPH_pred", "cohort"),
      pval = TRUE,
      conf.int = FALSE,
      xlim = c(0, 4000)
    )
  )
```

## KM plots {.tabset .tabset-pills}
```{r fig.width=10, fig.height=8, results="asis"}
survplots %>%
  iwalk(function(p, n) {
    cat(str_c("### ", n, "\n\n"))
    print(p)
    cat("\n\n")
  })
```

## Forest plot

We only consider samples predicted to be inflamed and desert.

```{r}
fits_to_df <- function(l, d) {
  imap_dfr(l,
    function(f, co) {
      n_class <- d %>%
        filter(cohort == co) %>%
        count(CD8IMMPH_pred) %>%
        pivot_wider(names_from = CD8IMMPH_pred, names_prefix = "n_", values_from = n, values_fill = 0)
      s <- summary(f)
      tibble(
        HR = s$conf.int[, "exp(coef)"],
        HR_lower = s$conf.int[, "lower .95"],
        HR_upper = s$conf.int[, "upper .95"],
        n = s$n,
        n_events = s$nevent
      ) %>%
        bind_cols(n_class)
    },
    .id = "cohort"
  )
}

fit_tcga <- function(iphe, d) {
  large_cohorts <- d %>%
    distinct(bcr_patient_barcode, cohort, CD8IMMPH_pred) %>%
    count(cohort, CD8IMMPH_pred) %>%
    pivot_wider(names_from = CD8IMMPH_pred, values_from = n, values_fill = 0) %>%
    filter(inflamed > 10, desert > 10)

  large_cohorts <- d %>%
    group_by(cohort) %>%
    filter(event == 1) %>%
    tally() %>%
    filter(n > 15) %>%
    inner_join(large_cohorts, by = "cohort")

  fits <- unique(large_cohorts$cohort) %>%
    set_names() %>%
    map(
      function(co) {
        survival::coxph(survival::Surv(time, event) ~ CD8IMMPH_pred == iphe,
          subset = cohort == co,
          data = d
        )
      }
    )
}

tcga_forest_plot <- function(d, iphe) {
  mod_df <- fit_tcga(iphe, d) %>%
    fits_to_df(d)
  m <- with(mod_df, meta::metagen(log(HR),
    lower = log(HR_lower), upper = log(HR_upper),
    studlab = str_glue("{cohort} (N={n})"), sm = "HR",
    data = mod_df
  ))
  meta::forest(m,
    xlim = c(0.5, 2), pooled.totals = TRUE, label.e = "events",
    leftcols = c("studlab", "n_inflamed", "n_desert", "n_events"),
    leftlabs = c("Study", "N inflamed", "N desert", "N events")
  )
}
```


```{r fig.width=12, fig.height=7}
d <- TCGA_surv_imm_phe %>%
  filter(CD8IMMPH_pred %in% c("inflamed", "desert"))

{
  svg(
    here::here("figures/fig6b_tcga_forest.svg"),
    width = 12,
    height = 7
  )
  tcga_forest_plot(d, "inflamed")
  dev.off()
}
tcga_forest_plot(d, "inflamed")
```

# Correlation with mutations
Here we are using missense & nonsense somatic mutations

```{r fig.width=8, fig.height=5}
library(RTCGA.mutations)
mut_data <- ls("package:RTCGA.mutations") %>%
  map(function(d) {
    get(d) %>%
      mutate(id = d) %>%
      filter(Variant_Classification %in% c("Missense_Mutation", "Nonsense_Mutation")) %>%
      filter(Mutation_Status == "Somatic") %>%
      group_by(id, bcr_patient_barcode) %>%
      summarize(n = n(), .groups = "drop")
  }) %>%
  bind_rows() %>%
  filter(str_starts(bcr_patient_barcode, "TCGA-")) %>%
  mutate(bcr_patient_barcode = str_replace(bcr_patient_barcode, "-[^-]{2}$", "")) %>%
  group_by(bcr_patient_barcode) %>%
  summarize(n = median(n))



mut_merged <- tcga_subset_pred %>%
  inner_join(mut_data, by = c("tcga.xml_bcr_patient_barcode" = "bcr_patient_barcode"))

mut_merged %>%
  group_by(cohort) %>%
  mutate(n_median = median(n)) %>%
  ungroup() %>%
  arrange(-n_median) %>%
  mutate(cohort = fct_inorder(cohort)) %>%
  ggplot() +
  aes(cohort, n, fill = CD8IMMPH_pred) +
  geom_boxplot() +
  scale_y_log10() +
  ylab("Number of nonsense and missense mutations") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = pheno_col)
```

# Association with LAMP5 and NKD1 expression levels
```{r "classify as low/high"}
gene_ids <- fData(tcga_subset) %>%
  filter(gene_name %in% c("LAMP5", "NKD1")) %>%
  {
    set_names(rownames(.), .$symbol)
  }

gene_median_expr <- map(
  gene_ids,
  ~ median(exprs(tcga_subset)[.x, ])
)

high_low <- function(symbol) {
  if_else(
    exprs(tcga_subset)[gene_ids[symbol], ] > gene_median_expr[symbol],
    str_c(symbol, "_HIGH"),
    str_c(symbol, "_LOW")
  ) %>%
    factor() %>%
    fct_rev()
}

pData(tcga_subset)$NKD1 <- high_low("NKD1")
pData(tcga_subset)$LAMP5 <- high_low("LAMP5")
```

```{r fig.height=8}
tcga_expression_summary <- pData(tcga_subset) %>%
  select(study, LAMP5, NKD1) %>%
  pivot_longer(-study,
    values_transform = function(x) str_replace(x, ".*_", ""),
    names_to = "gene",
    values_to = "status"
  ) %>%
  group_by(study, gene) %>%
  mutate(prop_high = sum(status == "HIGH") / n(), count = n()) %>%
  ungroup() %>%
  mutate(study = fct_reorder(study, -prop_high))

tcga_expression_summary %>%
  ggplot() +
  aes(study, fill = status) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = count, x = study, fill = NULL),
    data = distinct(tcga_expression_summary, study, count),
    y = 1.01, angle = 90, hjust = 0
  ) +
  ylim(0, 1.09) +
  facet_wrap(~gene, ncol = 1) +
  ylab("proportion")

ggsave(here::here("figures/sup_fig7b_tcga_lapm5_nkd1_status.svg"))
```

```{r}
pData(tcga_subset) %>%
  select(study, LAMP5, NKD1) %>%
  mutate(status = str_glue("{LAMP5} {NKD1}")) %>%
  group_by(study) %>%
  mutate(phigh = sum(status == "LAMP5_HIGH NKD1_HIGH") / n()) %>%
  ungroup() %>%
  arrange(-phigh) %>%
  mutate(study = factor(study) %>% fct_inorder()) %>%
  ggplot() +
  aes(study, fill = status) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = count, x = study, fill = NULL),
    data = distinct(tcga_expression_summary, study, count),
    y = 1.01, angle = 90, hjust = 0
  ) +
  ylim(0, 1.09) +
  ylab("proportion")
```

```{r message=FALSE, warning=FALSE}
tcga_subset_expr <- pData(tcga_subset) %>%
  select(tcga.xml_bcr_patient_barcode, LAMP5, NKD1, cohort = study) %>%
  mutate(across(c(LAMP5, NKD1), ~ str_replace(.x, ".*_", ""))) %>%
  mutate(across(c(LAMP5, NKD1), factor, levels = c("LOW", "HIGH")))

TCGA_surv_expr <- TCGA_surv %>%
  inner_join(tcga_subset_expr, by = c("bcr_patient_barcode" = "tcga.xml_bcr_patient_barcode")) %>%
  rename(time = times, event = patient.vital_status)

table(tcga_subset_expr$cohort, useNA = "always")


survplots <- unique(TCGA_surv_expr$cohort) %>%
  set_names() %>%
  map(
    ~ RTCGA::kmTCGA(
      filter(TCGA_surv_expr, cohort == .x),
      times = "time",
      status = "event",
      explanatory.names = c("LAMP5", "NKD1", "cohort"),
      pval = TRUE,
      conf.int = FALSE,
      xlim = c(0, 4000)
    )
  )
```

## KM plots {.tabset .tabset-pills}
```{r fig.width=10, fig.height=8, results="asis"}
survplots %>%
  iwalk(function(p, n) {
    cat(str_c("### ", n, "\n\n"))
    print(p)
    cat("\n\n")
  })
```

```{r}
fit_tcga_expr <- function(d, gene) {
  large_cohorts <- d %>%
    distinct(bcr_patient_barcode, cohort, LAMP5, NKD1) %>%
    group_by(cohort) %>%
    filter(
      sum(.data[[gene]] == "HIGH") > 10,
      sum(.data[[gene]] == "LOW") > 10
    ) %>%
    distinct(cohort)

  large_cohorts <- d %>%
    group_by(cohort) %>%
    filter(event == 1) %>%
    tally() %>%
    filter(n > 15) %>%
    inner_join(large_cohorts, by = "cohort")

  unique(large_cohorts$cohort) %>%
    set_names() %>%
    map(
      function(co) {
        survival::coxph(
          as.formula(str_glue("survival::Surv(time, event) ~ {gene}")),
          subset = cohort == co,
          data = d
        )
      }
    )
}

fits_expr_to_df <- function(l, d, gene) {
  imap_dfr(l,
    function(f, co) {
      n_high <- sum(d[d$cohort == co, gene] == "HIGH")
      s <- summary(f)
      tibble(
        coef = str_replace(rownames(s$conf.int), "HIGH$", " HIGH"),
        HR = s$conf.int[, "exp(coef)"],
        HR_lower = s$conf.int[, "lower .95"],
        HR_upper = s$conf.int[, "upper .95"],
        n = s$n,
        n_high = n_high,
        n_events = s$nevent
      )
    },
    .id = "cohort"
  )
}

tcga_forest_expr_plot <- function(d, gene) {
  mod_df <- fit_tcga_expr(d, gene) %>%
    fits_expr_to_df(d, gene)
  m <- with(mod_df, meta::metagen(log(HR),
    lower = log(HR_lower), upper = log(HR_upper),
    studlab = str_glue("{cohort}"), sm = "HR",
    data = mod_df
  ))
  meta::forest(m,
    xlim = c(0.5, 2), pooled.totals = TRUE, label.e = "events",
    leftcols = c("studlab", "n", "n_high", "n_events"),
    leftlabs = c("Study", "N", "N high", "N events")
  )
  grid::grid.text(str_glue("{gene} high"), .5, .9, gp = grid::gpar(cex = 1.5))
}
```

## Forest plot

```{r fig.width=10, fig.height=7.5}
{
  fn <- "figures/sup_fig7_tcga_lamp5_nkd1_forest_%02d.svg"
  svglite::svglite(
    here::here(fn),
    width = 12,
    height = 7
  )
  tcga_forest_expr_plot(TCGA_surv_expr, "LAMP5")
  tcga_forest_expr_plot(TCGA_surv_expr, "NKD1")
  dev.off()

  # replace [1.23; 4.56] to [1.23-4.56]
  cmd <- r"(sed -i 's/\(\[[0-9]*\(\.[0-9]*\)\?\); \([0-9]*\(\.[0-9]*\)\?\)\]/\1–\3]/g')"
  system(paste(cmd, here::here(sprintf(fn, 1))))
  system(paste(cmd, here::here(sprintf(fn, 2))))
}
tcga_forest_expr_plot(TCGA_surv_expr, "LAMP5")
tcga_forest_expr_plot(TCGA_surv_expr, "NKD1")
```

<details>
<summary>Session info</summary>
```{r}
devtools::session_info()
```
</details>
